Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-16 numerical-mhd-ct.tex
Adding AMR scaling section with an associated table.  I still need to add some references in this section.  Fixing some undefined and multiple labels and making the UVB table wider and spaced nicely.

This MHD method was developed by \citet{Collins10} in order to study Population
III stars \citep{Xu08}, molecular clouds \citep{collins12}, and galaxy clusters
\citep{Xu12, Skillman13}.  A full description and suite of test problems can be
found in \citet{Collins10}

@@

Dual energy formalism has also been incorporated in two methods, one internal
energy, similar to that described in Section \ref{sec.hydro.ppm}, and one that
uses entropy.  The additional equations are
\begin{eqnarray}
\end{eqnarray}
\begin{eqnarray}
  \frac{\partial S}{\partial t} +\frac{1}{a} \nabla \cdot (S \vecv) & = &  -
  \frac{ 3 (\gamma - 1) \dot{a}}{a} S \\
  \frac{\partial \rho e}{\partial t} + \frac{1}{a} \nabla \cdot (\rho e \vecv) &
  = & -\frac{ 3(\gamma -1 ) \dot{a}}{a} + \frac{p}{a} \nabla \cdot \vecv,
\end{eqnarray}
where $e$ is the internal energy and $S=p/\rho^{\gamma-1}$ is the entropy.  If
the ratio of thermal to kinetic energy, $\eta$, is below a certain value, the
alternative energy is used.  As in \citet{Li08}, we use $\eta=0.008$.

>>>>

This MHD method was developed by \citet{Collins10} in order to study Population
III stars \citep{Xu08}, molecular clouds \citep{Collins12}, and galaxy clusters
\citep{Xu12, Skillman13}.  A full description and suite of test problems can be
found in \citet{Collins10}

@@

Dual energy formalism has also been incorporated in two methods, one internal
energy, similar to that described in Section \ref{sec.hydro.ppm}, and one that
uses entropy.  The additional equations are
\begin{eqnarray}
\end{eqnarray}
\begin{eqnarray}
  \frac{\partial S}{\partial t} +\frac{1}{a} \nabla \cdot (S \vecv) & = &  -
  \frac{ 3 (\gamma - 1) \dot{a}}{a} S \\
  \frac{\partial \rho e}{\partial t} + \frac{1}{a} \nabla \cdot (\rho e \vecv) &
  = & -\frac{ 3(\gamma -1 ) \dot{a}}{a} + \frac{p}{a} \nabla \cdot \vecv,
\end{eqnarray}
where $e$ is the internal energy and $S=p/\rho^{\gamma-1}$ is the entropy.  If
the ratio of thermal to kinetic energy, $\eta$, is below a certain value, the
alternative energy is used.  As in \citet{Li08a}, we use $\eta=0.008$.

### 2013-05-16 parallel.tex
Adding AMR scaling section with an associated table.  I still need to add some references in this section.  Fixing some undefined and multiple labels and making the UVB table wider and spaced nicely.

\subsubsection{Unigrid scaling}

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/enzo_unigrid_weak_scaling.eps}
\caption{\enzo\ weak scaling performance for a set of Lyman alpha
forest cosmology simulations with constant comoving spatial resolution
per grid cell, showing cell updates per second per computational core
plotted as a function of the number of root grid tiles of dimension
$128^3$ (R) in each dimension.  The number of MPI tasks is N$ = R^3$,
so R$ = 16$ on this plot corresponds to a $2048^3$ computational mesh
running on 4096 MPI tasks.  This plot goes from R$ = 2$ (8 MPI tasks)
to R$ = 24$ (13,824 MPI tasks) on two supercomputers -- NICS Kraken
and ORNL Jaguar when they were Cray XT4 systems -- and using 1, 2 or 4
MPI processes per node, where each compute node contained a single
quad-core AMD Opteron CPU having a speed of 2.1 GHz on Jaguar and 2.3
GHz on Kraken.}
\label{fig.uniscale}
\end{center}
\end{figure}

@@

\subsubsection{AMR scaling \red{(John)}}

\begin{figure}
\begin{center}
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_levels.eps}
\hfill
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_routines.eps}
\end{center}
\caption{\emph{Left:} Strong scaling test of a 512$^3$ AMR
  cosmological calculation.  The root grid scaling is not
  representative of the true strong scaling of Enzo because the root
  grid tiles are not repartitioned when a simulation is restarted with
  additional cores.  The weak scaling test in Figure
  \ref{fig.uniscale} is more representative of the scaling on the root
  level.  The performance in the refined levels show good strong
  scaling up to 4096 cores.  \emph{Right:} Time spent in the major
  routines in the same AMR calculation.}
\label{fig:amr_hierarchy}
\end{figure}

% Greg to work on this, scaling data from his runs, or maybe from John Wise

>>>>

\subsubsection{Unigrid scaling}
\label{sec:weak_scaling}

\begin{figure}
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/enzo_unigrid_weak_scaling.eps}
\caption{\enzo\ weak scaling performance for a set of Lyman alpha
forest cosmology simulations with constant comoving spatial resolution
per grid cell, showing cell updates per second per computational core
plotted as a function of the number of root grid tiles of dimension
$128^3$ (R) in each dimension.  The number of MPI tasks is N$ = R^3$,
so R$ = 16$ on this plot corresponds to a $2048^3$ computational mesh
running on 4,096 MPI tasks.  This plot goes from R$ = 2$ (8 MPI tasks)
to R$ = 24$ (13,824 MPI tasks) on two supercomputers -- NICS Kraken
and ORNL Jaguar when they were Cray XT4 systems -- and using 1, 2 or 4
MPI processes per node, where each compute node contained a single
quad-core AMD Opteron CPU having a speed of 2.1 GHz on Jaguar and 2.3
GHz on Kraken.}
\label{fig.uniscale}
\end{center}
\end{figure}

@@

\subsubsection{AMR scaling}

\begin{figure}
\begin{center}
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_levels.eps}
\hfill
\includegraphics[width=0.48\textwidth]{figures/strong_scaling_routines.eps}
\end{center}
\caption{\emph{Left:} Strong scaling test of a 512$^3$ AMR
  cosmological calculation.  The root grid scaling is not
  representative of the true strong scaling of Enzo because the root
  grid tiles are not repartitioned when a simulation is restarted with
  additional cores.  The weak scaling test in Figure
  \ref{fig.uniscale} is more representative of the scaling on the root
  level.  The performance in the refined levels show good strong
  scaling up to 2,048 cores.  \emph{Right:} Time spent in
  representative major routines in the same AMR calculation.}
\label{fig:strong_scaling}
\end{figure}

Many astrophysical problems, such as cosmological galaxy formation
\citep{}, high-resolution disk galaxy simulations \citep{},
high-redshift \citep{} and present-day star formation \citep{},
involve multi-scale physics that span several orders of magnitude in
both space and time.  In these situations, using Enzo in its adaptive
mesh refinement mode is beneficial.  Because the adaptive grid
hierarchy is dynamic and not static, grid boundaries and, thus,
communication patterns can be unpredictable, hindering good strong
scaling at high core counts.

Figure \ref{fig:strong_scaling} shows strong scaling results from a
single 50 Mpc/h cosmology simulation run on $N_{\rm core}$ compute
cores, ranging from 128 to 16,384 at power-of-two intervals.  It was
run on the NICS Kraken XT5 supercomputer that has two AMD Opteron
hexa-core processors with 16 GB of memory per compute node.  The
simulations that utilized 128, 256, and 512 cores were executed on 128
nodes because of the memory requirements.  The higher core count jobs
were run with 8 cores per node.  This simulation would not run with 12
cores per node because of the memory overhead associated with the grid
hierarchy being duplicated on each MPI process.  However, this
overhead is greatly diminished if a hybrid-parallel (MPI + OpenMP)
approach is used.

This simulation includes dark matter dynamics, hydrodynamics using the
piecewise parabolic method, six-species non-equilibrium chemistry and
radiative cooling, and a uniform metagalactic ultraviolet background.
The simulation uses the space-filling curve method for load balancing
the AMR grids.  It has a $512^3$ root grid that is divided into 512
tiles, and 6 additional AMR levels are used.  We perform these scaling
tests when the simulation reaches $z=4$.  The results shown in Figure
\ref{fig:strong_scaling} come from a single root-level timestep
$\Delta t = 2.1\, \textrm{Myr}$.  At this time, there are $3.04 \times
10^5$ AMR grids, $9.03 \times 10^8$ ($\sim 967^3$) computational
cells, and $1.34 \times 10^8$ dark matter particles in total.  The
breakdown of the number of AMR grids, cells, timesteps, and number of
cell updates on each level is shown in Table \ref{tab:amr_scale}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table*}
  \begin{center}
  \caption{Strong scaling test computational details}
  \begin{tabular*}{0.9\textwidth}{@{\extracolsep{\fill}}c c c c c}
    \tableline\tableline
    {Level} & {$N_{\rm grid}$} & {$N_{\rm cells}$} & {$N_{\rm up}$} &
    {$N_{\rm timesteps}$}\\
    \tableline
    0 & 512 & $1.34 \times 10^8$ & $1.34 \times 10^8$ & 1\\
    1 & 61,868 & $4.01 \times 10^8$ & $4.01 \times 10^8$ & 1\\
    2 & 91,182 & $1.99 \times 10^8$ & $5.96 \times 10^8$ & 3\\
    3 & 59,932 & $7.62 \times 10^7$ & $5.34 \times 10^8$ & 7\\
    4 & 40,700 & $3.32 \times 10^7$ & $5.65 \times 10^8$ & 17\\
    5 & 28,346 & $2.76 \times 10^7$ & $1.36 \times 10^9$ & 49\\
    6 & 19,771 & $2.80 \times 10^7$ & $5.25 \times 10^9$ & 187\\
    \tableline
    Total & 302,311 & $9.03 \times 10^8$ & $8.83 \times 10^9$ & --\\
  \end{tabular*}
  \parbox[t]{0.9\textwidth}{\textbf{Note.} --- Data shown at $z=4$ for
    a root grid timestep of 2.1~Myr.  Col. (1): AMR Level. Col. (2):
    Number of grids. Col. (3): Number of computational
    cells. Col. (4): Number of cell updates. Col. (5): Number of
    timesteps.}
  \label{tab:amr_scale}
  \end{center}
\end{table*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The left panel in Figure \ref{fig:strong_scaling} shows the
computational and communication time spent on each level.  In the AMR
levels, there exists good strong scaling up to 2,048 cores, and
marginal speed-ups are found at 4,096 cores.  On the root-grid level,
there exists good scaling up to 1,024 cores, but the performance
dramatically decreases at higher core counts.  This occurs because the
root grid is not re-partitioned into $N_{\rm core}$ tiles when the
simulation is restarted with a different core count.  This feature can
be easily implemented and is planned in the next major release of
Enzo, where scaling results would be similar to the weak scaling shown
in \S\ref{sec:weak_scaling}.  The right panel in Figure
\ref{fig:strong_scaling} shows the time spent in some representative
major routines in Enzo.  The local physics routines, for example
\texttt{SolveHydroEquations}, exhibit perfect strong scaling because
they involve no communication.  By investigating the scaling behavior
in each routine, it is clear that the communication in the
\texttt{SetBoundaryConditions}, \texttt{SolveForPotential} (multi-grid
solver in AMR levels), and \texttt{RebuildHierarchy} are responsible
for the lack of strong scaling at $N_{\rm core} \ga$~4,096 in this
particular simulation.  The transpose of the root grid tiles are
responsible for the performance decrease because it is not optimized
for situations where the number of tiles is greater than the number of
MPI processes.  These results are directly applicable to simulations
with similar computational demands.  In simulations with less
computational cells, strong scaling will cease at a smaller core count
because the CPUs will become data-starved faster, and the opposite
occurs with larger simulations.
